
\title{Intelligent Clustering in Wireless Sensor Networks}%
\author{Robert William Lee Guderian}

\degreemonth{July} % month final submission occurs.
\degreeyear{2012}%
\degree{Master of Science}%
\department{Computer Science}%
\advisor{Dr.\ Rasit Eskicioglu} %


%\copyrightpage
% Insert a blank page for two-sided
%  \newpage
%  \thispagestyle{empty}
%  \hbox{}
%  \newpage


\maketitle

\begin{abstract}

	Wireless Sensor Networks (WSNs) are networks of small devices, called motes,
    	designed to monitor resources and report to a server. Motes are 
    	battery-powered and have very little memory to store data. To conserve power, 
    	the motes usually form clusters to coordinate their activities. In 
    	heterogeneous WSNs, the motes have different resources available to them. For 
    	example, some motes might have more powerful radios, or larger power supplies. 
    By exploiting heterogeneity within a WSN can allow 
    	the network to stay active for longer periods of time.

    	In WSNs, the communications between motes draw the most power. By choosing 
    	better clusterheads in the clusters to control and route 
    	messages, all motes in the network will have longer lifespans. By leveraging 
    	heterogeneity to select better clusterheads, I have developed
    	Heterogeneous Clustering Control Protocol (HCCP). HCCP is designed to be 
    	highly robust to change and to fully utilize the resources that are currently 
    	available.

    


\end{abstract}

\newpage
\tableofcontents

\addcontentsline{toc}{section}{Table of Contents}
%comment these out if you don't want a detailed list of figures and tables!
\listoffigures
\listoftables

\begin{acknowledgments}
I would like to begin by thanking my Dr.\ Rasit Eskicioglu who has been my
long-suffering advisor to whom I will forever be grateful. I'd like to thank
my committee, Dr.\ Peter Graham and Dr.\ Bob McLeod for all their input
and patience. I'd finally like to thank my biggest fan, my inspiration, my
editor, my best friend and wife, Becky.
\end{acknowledgments}

%
\dedication
\begin{quote}
\hsp \em \raggedleft Dedicated to Becky for always believing in me.
\end{quote}

%\newpage
\cleardoublepage
\thispagestyle{empty}
\startarabicpagination

%%% end
